---
title: "Diversity tests"
author: "Alexander Piper"
date: "`r Sys.Date()`"
output:
  
  html_document:
    highlighter: null
    theme: "flatly"
    code_download: true
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: true
    df_print: paged    
  pdf_document: default
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE, error = FALSE)
```


# Load packages

```{r load packages}

#Set required packages
.cran_packages <- c("tidyverse",
                    "tidymodels",
                    "patchwork", 
                    "vegan", 
                    "ape", 
                    "RColorBrewer",
                    "picante",
                    "phytools",
                    "ggrepel",
                    "devtools")
.bioc_packages <- c("dada2",
                    "phyloseq", 
                    "ggtree",
                    "philr",
                    "ALDEx2")

.inst <- .cran_packages %in% installed.packages()
if(any(!.inst)) {
   install.packages(.cran_packages[!.inst])
}
.inst <- .bioc_packages %in% installed.packages()
if(any(!.inst)) {
  if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
  BiocManager::install(.bioc_packages[!.inst], ask = F)
}

sapply(c(.cran_packages,.bioc_packages), require, character.only = TRUE)

# Github packages
devtools::install_github("alexpiper/taxreturn")
devtools::install_github("alexpiper/seqateurs")
devtools::install_github("mikemc/speedyseq")

library(speedyseq)
library(taxreturn)
library(seqateurs)
```

# Read in phyloseq object

```{r phyloseq}
ps <- readRDS("output/rds/ps.rds")
```


# Alpha diversity (Unrarefied)

```{r alpha div}
dir.create("output/alpha")

# Get a histogram of taxon sums
taxa_sums(ps) %>%
  as_tibble(rownames = "OTU") %>%
  filter(value > 0) %>%
  ggplot(aes(x = value)) +
  geom_histogram() 

# Get Alpha diversity measures
richness <- phyloseq::estimate_richness(ps, measures=c("Chao1", "Shannon", "Simpson")) %>% 
  rownames_to_column("sample_id")

# Calculate Faith's phylogenetic diversity
sespd <- picante::ses.pd(as(phyloseq::otu_table(ps), "matrix"),  phyloseq::phy_tree(ps), null.model = "taxa.labels", include.root = F, runs = 9)

# Join together
div_table <- sespd %>%
  rownames_to_column("sample_id") %>%
  dplyr::select(sample_id, ntaxa, pd.obs) %>%
  dplyr::rename(pd = pd.obs, alpha = ntaxa) %>%
  left_join(richness, by="sample_id") %>%
  left_join(sample_data(ps) %>% 
              as("data.frame"),
            by = "sample_id") 


# Summarise mean diversity
div_table %>%
  summarise(across(c("alpha", "pd", "Chao1", "se.chao1", "Shannon", "Simpson"), mean, na.rm=TRUE))

# Difference in alpha diversity between a factor
# In this case we are going to use sample_name as that factor
report::report(aov(alpha ~sample_name, data=div_table))
broom::tidy(TukeyHSD(aov(alpha ~sample_name, data=div_table)))

report::report(aov(Shannon ~sample_name, data=div_table))
broom::tidy(TukeyHSD(aov(Shannon ~sample_name, data=div_table)))

report::report(aov(pd ~sample_name, data=div_table))
broom::tidy(TukeyHSD(aov(pd ~sample_name, data=div_table)))

# plot differences in alpha diversity
gg.alpha <- div_table %>% 
  pivot_longer(c("alpha", "pd", "Chao1", "se.chao1", "Shannon", "Simpson"),
               names_to="metric",
               values_to="value") %>%
  ggplot(aes(x=fcid, y=value, colour=fcid)) +
  geom_jitter(size=2, alpha=0.6, width=0.2, height=0)+
  geom_boxplot(fill=NA)+
  #coord_flip()+
  scale_colour_brewer(palette="Paired") +
  facet_wrap(.~metric,ncol=1, scales="free")

gg.alpha

# Write out results as a CSV
alpha_results <- bind_rows(
  broom::tidy(TukeyHSD(aov(alpha ~sample_name, data=div_table))) %>% mutate(metric = "richness"),
  broom::tidy(TukeyHSD(aov(Shannon ~sample_name, data=div_table))) %>% mutate(metric = "shannon"),
  broom::tidy(TukeyHSD(aov(pd ~sample_name, data=div_table))) %>% mutate(metric = "phylogenetic")
)

write_csv(alpha_results, "output/alpha/alpha_anova.csv")
```


# Alpha diversity (Rarefied)

```{r alpha div rarefied}
dir.create("output/alpha_rarefied")

# Get a histogram of taxon sums
taxa_sums(ps) %>%
  as_tibble(rownames = "OTU") %>%
  filter(value > 0) %>%
  ggplot(aes(x = value)) +
  geom_histogram() 

# Get a histogram of sample sums to choose rarefaction depth
sample_sums(ps) %>%
  as_tibble(rownames = "sample") %>%
  filter(value > 0) %>%
  ggplot(aes(x = value)) +
  geom_histogram() +
  scale_x_continuous(labels = scales::label_number_si()) +
  labs(x = "Sequence Reads",
       y = "Number of samples")

# Choose a depth to rarefy to (everything under this will be discarded)
raredepth <- 100000


# Plot a rarefaction curve to see which samples would be lost at this depth
rare <- otu_table(ps) %>%
  as("matrix") %>%
  rarecurve(step=10000) %>% # Change this to be higher if you have heaps of reads to make it faster
  purrr::map_dfr(., function(x){
  b <- as.data.frame(x)
  b <- data.frame(OTU = b[,1], count = rownames(b))
  b$count <- as.numeric(gsub("N", "",  b$count))
  nm <- names(attr(x, "Subsample"))
  b$sample_id <- nm[!nm==""]
  return(b)
  })%>%
  left_join(sample_data(ps)%>%
    as("matrix") %>%
    as_tibble() %>%
      dplyr::select(sample_id, fcid) %>%
      distinct())

gg.rare <- rare %>%
  ggplot() +
  geom_line(aes(x = count, y = OTU, group=sample_id), alpha=0.5)+
  geom_point(data = rare %>% 
               group_by(sample_id) %>% 
               top_n(1, count),
             aes(x = count, y = OTU, colour=(count > raredepth))) +
  scale_x_continuous(labels =  scales::label_number_si()) +
  geom_vline(xintercept=raredepth, linetype="dashed") +
  facet_wrap(fcid~., scales="free", ncol=1)+
  theme(legend.position = "bottom")+
  labs(x = "Sequence reads",
       y = "Observed ASV's",
       colour = "Sample retained?") 

gg.rare

# Rarefy (make sure to set the rngseed for reproducability)
ps_rare <- rarefy_even_depth(ps, sample.size = raredepth,
  rngseed = 532251, replace = TRUE, trimOTUs = TRUE, verbose = TRUE)

rarestats <- speedyseq::psmelt(ps_rare)
  
# Get Alpha diversity measures
richness_rare <- phyloseq::estimate_richness(ps_rare, measures=c("Chao1", "Shannon", "Simpson")) %>% 
  rownames_to_column("sample_id")

# Calculate Faith's phylogenetic diversity
sespd_rare <- picante::ses.pd(as(phyloseq::otu_table(ps_rare), "matrix"),  phyloseq::phy_tree(ps_rare), null.model = "taxa.labels", include.root = F, runs = 9)

# Join together
div_table_rare <- sespd_rare %>%
  rownames_to_column("sample_id") %>%
  dplyr::select(sample_id, ntaxa, pd.obs) %>%
  dplyr::rename(pd = pd.obs, alpha = ntaxa) %>%
  left_join(richness_rare, by="sample_id") %>%
  left_join(sample_data(ps_rare) %>% 
              as("data.frame"),
            by = "sample_id") 


# Summarise mean diversity
div_table_rare %>%
  summarise(across(c("alpha", "pd", "Chao1", "se.chao1", "Shannon", "Simpson"), mean, na.rm=TRUE))

# Difference in alpha diversity between a factor
# In this case we are going to use sample_name as that factor
report::report(aov(alpha ~sample_name, data=div_table_rare))
broom::tidy(TukeyHSD(aov(alpha ~sample_name, data=div_table_rare)))

report::report(aov(Shannon ~sample_name, data=div_table_rare))
broom::tidy(TukeyHSD(aov(Shannon ~sample_name, data=div_table_rare)))

report::report(aov(pd ~sample_name, data=div_table_rare))
broom::tidy(TukeyHSD(aov(pd ~sample_name, data=div_table_rare)))

# plot differences in alpha diversity
gg.alpha_rare <- div_table_rare %>% 
  pivot_longer(c("alpha", "pd", "Chao1", "se.chao1", "Shannon", "Simpson"),
               names_to="metric",
               values_to="value") %>%
  ggplot(aes(x=fcid, y=value, colour=fcid)) +
  geom_jitter(size=2, alpha=0.6, width=0.2, height=0)+
  geom_boxplot(fill=NA)+
  #coord_flip()+
  scale_colour_brewer(palette="Paired") +
  facet_wrap(.~metric,ncol=1, scales="free")

gg.alpha

# Write out results as a CSV
alpha_results_rare <- bind_rows(
  broom::tidy(TukeyHSD(aov(alpha ~sample_name, data=div_table_rare))) %>% mutate(metric = "richness"),
  broom::tidy(TukeyHSD(aov(Shannon ~sample_name, data=div_table_rare))) %>% mutate(metric = "shannon"),
  broom::tidy(TukeyHSD(aov(pd ~sample_name, data=div_table_rare))) %>% mutate(metric = "phylogenetic")
)

write_csv(alpha_results, "output/alpha_rarefied/alpha_anova.csv")
```

# Beta diversity

First calculate the following beta diversity between samples

* Jaccard - A presence absense metric
* Bray-curtis - A abundance weighted metric
* Aitchison - A compositional and abundance weighted metric
* Philr - A compositional, abundance and phylogenetic weighted metric
* Unifrac - A phylogenetic presence / absence metric
* Weighted Unifrac - A abundance and phylogenetic weighted metric

```{r Beta distances}
# Get OTU tables
otutab <- otu_table(ps) %>%
  as("matrix")

#Impute zeroes for the compositional distances
otutab_n0 <- as.matrix(zCompositions::cmultRepl(otutab, method="CZM", output="p-counts"))

#Root phylogenetic tree
phy_tree(ps) <- multi2di(phy_tree(ps))
phy_tree(ps) <- makeNodeLabel(phy_tree(ps), method="number", prefix='n')
name.balance(phy_tree(ps), tax_table(ps), 'n1') #Get root

#Calculate different distance metrics
metrics <- c("Bray", "Jaccard", "Aitchison", "Philr", "Unifrac", "WUnifrac")
distlist <- vector("list", length=length(metrics))
names(distlist) <- metrics

distlist$Jaccard <- as.matrix(vegdist(otutab, method="jac",binary = T))
distlist$Bray <- as.matrix(vegdist(otutab, method="bray"))
distlist$Aitchison <- as.matrix(vegdist(CoDaSeq::codaSeq.clr(otutab_n0), method="euclidean"))
distlist$Philr <- as.matrix(vegdist(philr::philr(otutab_n0, phy_tree(ps),
                                                part.weights='enorm.x.gm.counts',
                                                ilr.weights='blw.sqrt'), method="euclidean"))
distlist$Unifrac <- as.matrix(phyloseq::UniFrac(ps, weighted=FALSE, parallel = TRUE))
distlist$WUnifrac <- as.matrix(phyloseq::UniFrac(ps, weighted=TRUE, parallel = TRUE))
```

# Adonis and betadisper

Now we will conduct PERMANOVA (adonis), and PERMDISP tests

```{r betatest}
# Adonis test
metadata <- sample_data(ps) %>%
  as("data.frame")

# Test difference in beta diversity ny sample (Change this to whatever factor your interested in)
adonis_results <- distlist %>%
  purrr::map(function(x) {
    bind_rows(
    broom::tidy(adonis(x~sample_name, method="euclidean", data=metadata)$aov.tab) 
    )
})  %>%
  bind_rows(.id="dist")

# Check homogeneity with betadisper
betadisper_results <- distlist %>%
  purrr::map(function(x) {
    y <- as.dist(x[metadata$sample_id, metadata$sample_id])
  bind_rows(
    as.data.frame(permutest(vegan::betadisper(y, metadata$sample_name))$tab) %>%
      #dplyr::slice(1) %>% 
      mutate(term="sample_name")
  )
})  %>%
  bind_rows(.id="dist")

dir.create("output/beta")
write_csv(adonis_results, "output/beta/adonis.csv")
write_csv(betadisper_results, "output/beta/adonis.csv")
```

# PCA plots

Finally we can plot a PCA plot of how things cluster with a different beta diversity metric

```{r pca plots}
# Get desired distance to make the plot (here i am using philr)
philr.dist <- as.dist(distlist$Philr)

#Philr PCOA
philr.pcoa <- ordinate(ps, 'PCoA', distance=philr.dist)

gg.pca <- plot_ordination(ps, philr.pcoa, color = "sample_name") +
  geom_point(size=3, alpha=0.8) +
  geom_hline(yintercept = 0, linetype=2, alpha=0.5) +  
  geom_vline(xintercept = 0, linetype=2, alpha=0.5) +
  #coord_fixed() +
  theme(legend.position = "none") +
  labs(colour = "Sample Type")

gg.pca
```


```{details, echo = FALSE, details.summary = 'Reproducability receipt'}
# datetime
Sys.time()
#repository
git2r::repository()
sessioninfo::session_info()
```